Covariates analysis of MSBB RNASeq data

### 2016-05-24 code/Rmd/MSSM_RNASeq.Rmd
Added section to store raw log cpm values to synapse

## Load required libraries
library(CovariateAnalysis)
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

@@

# Arrange expression and covariates
COUNT = COUNT[, rownames(COVARIATES)]
```
Covariate correlation
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.3, row.width=0.15)
```

@@

# Mean Plaque
p[[4]] = ggplot(METADATA, aes(x = Dx, y = as.numeric(NP.1), fill = BrainRegion)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('NP.1') + theme(legend.position = 'top')

@@

Remove genes that have less than 1 cpm counts in at least 50% of samples.
```{r cpmnormalisation, cache=TRUE}
PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT,MIN_GENE_CPM=1, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
```
### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T)
```
### Outlier analysis
Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T)

@@

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);
tmp = gsub('\\..*','',cols)
WGCNA::plotDendroAndColors(tree, 
                           colors = tmp, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
dev.off()
```
### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
COVARIATES.tmp = COVARIATES
COVARIATES.tmp[, FactorCovariates] = sapply(COVARIATES.tmp[, FactorCovariates], as.numeric)
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, COVARIATES.tmp,'NULL design(voom-normalized)', isKeyPlot=TRUE, MIN_PVE_PCT_PC = 1)
```
Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

@@

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES.tmp[, residCovars, drop=F], 'all adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

@@

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T)

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = rownames(RESIDUAL.GENE_EXPRESSION),
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id"),
                                      Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = as.data.frame(summary(COVARIATES$BrainRegion.Dx))
colnames(tmp) = 'No. of samples'
kable(tmp)
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=18}
# Fit contrast
contrast = makeContrasts(contrasts=c("BrainRegion.DxBM_10.ND-BrainRegion.DxBM_10.MD", 
                                     "BrainRegion.DxBM_10.ND-BrainRegion.DxBM_10.SD", 
                                     "BrainRegion.DxBM_10.MD-BrainRegion.DxBM_10.SD",
                                     "BrainRegion.DxBM_22.ND-BrainRegion.DxBM_22.MD", 
                                     "BrainRegion.DxBM_22.ND-BrainRegion.DxBM_22.SD", 
                                     "BrainRegion.DxBM_22.MD-BrainRegion.DxBM_22.SD",
                                     "BrainRegion.DxBM_36.ND-BrainRegion.DxBM_36.MD", 
                                     "BrainRegion.DxBM_36.ND-BrainRegion.DxBM_36.SD", 
                                     "BrainRegion.DxBM_36.MD-BrainRegion.DxBM_36.SD",
                                     "BrainRegion.DxBM_44.ND-BrainRegion.DxBM_44.MD", 
                                     "BrainRegion.DxBM_44.ND-BrainRegion.DxBM_44.SD", 
                                     "BrainRegion.DxBM_44.MD-BrainRegion.DxBM_44.SD"), 
                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT = eBayes(FIT)

# Get differnetial expression
DE = list()
p = list()
for (i in 1:dim(contrast)[2]){
  DE[[i]] = topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id")
  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
  p[[i]] <- p[[i]] + ggtitle(gsub('BrainRegion.Dx','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red')
}
multiplot(plotlist=p, cols = 3)
names(DE) = gsub('BrainRegion.Dx','',colnames(contrast))

@@

# Expression
expr = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'EnsembleIDs')
write.table(expr, file='expression.tsv', sep = '\t', row.names=F, quote=F)
obj = File('expression.tsv', 
           name = 'Voom Adjusted Weighted Residual Expression (BrainRegion.Dx added)', 
           parentId = CODE$properties$id)
expr_obj = synStore(obj,
               used = ALL_USED_IDs, 
               activityName = activityName, 
               executed = thisFile, 
               activityDescription = activityDescription)

>>>>

## Load required libraries
library(CovariateAnalysis) # Get the utilities package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

@@

# Arrange expression and covariates
COUNT = COUNT[, rownames(COVARIATES)]
```
Covariate correlation
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.5, row.width=0.1)
```

@@

# NP.1
p[[4]] = ggplot(METADATA, aes(x = Dx, y = as.numeric(NP.1), fill = BrainRegion)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('NP.1') + theme(legend.position = 'top')

@@

Remove genes that have less than 1 cpm counts in at least 50% of samples.
```{r cpmnormalisation, cache=TRUE}
PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT,MIN_GENE_CPM=1, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
```
### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T)
```
### Outlier analysis
Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)

@@

# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,FactorCovariates]))
dev.off()
```
### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, COVARIATES,'NULL design(voom-normalized)', isKeyPlot=TRUE, MIN_PVE_PCT_PC = 1)
```
Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

@@

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F], 'all adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

@@

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale. = T, center = T)

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = rownames(RESIDUAL.GENE_EXPRESSION),
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id"),
                                      Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
VOOM.ADJUSTED.GENE_EXPRESSION$E <- left_join(rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, "ensembl_gene_id"),
                                             Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = as.data.frame(summary(COVARIATES$BrainRegion.Dx))
colnames(tmp) = 'No. of samples'
kable(tmp)
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=18}
# Fit contrast
contrast = makeContrasts(contrasts=c("BrainRegion.DxBM_10.ND-BrainRegion.DxBM_10.MD", 
                                     "BrainRegion.DxBM_10.ND-BrainRegion.DxBM_10.SD", 
                                     "BrainRegion.DxBM_10.MD-BrainRegion.DxBM_10.SD",
                                     "BrainRegion.DxBM_22.ND-BrainRegion.DxBM_22.MD", 
                                     "BrainRegion.DxBM_22.ND-BrainRegion.DxBM_22.SD", 
                                     "BrainRegion.DxBM_22.MD-BrainRegion.DxBM_22.SD",
                                     "BrainRegion.DxBM_36.ND-BrainRegion.DxBM_36.MD", 
                                     "BrainRegion.DxBM_36.ND-BrainRegion.DxBM_36.SD", 
                                     "BrainRegion.DxBM_36.MD-BrainRegion.DxBM_36.SD",
                                     "BrainRegion.DxBM_44.ND-BrainRegion.DxBM_44.MD", 
                                     "BrainRegion.DxBM_44.ND-BrainRegion.DxBM_44.SD", 
                                     "BrainRegion.DxBM_44.MD-BrainRegion.DxBM_44.SD"), 
                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT = eBayes(FIT)

# Get differnetial expression
DE = list()
p = list()
for (i in 1:dim(contrast)[2]){
  DE[[i]] = topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id")
  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
  p[[i]] <- p[[i]] + ggtitle(gsub('BrainRegion.Dx','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red')
}
multiplot(plotlist=p, cols = 4)
names(DE) = gsub('BrainRegion.Dx','',colnames(contrast))

@@

# Expression
write.table(RESIDUAL.GENE_EXPRESSION, file='expression.tsv', sep = '\t', row.names=F, quote=F)
obj = File('expression.tsv', 
           name = 'Voom Adjusted Weighted Residual Expression (BrainRegion.Dx added)', 
           parentId = CODE$properties$id)
expr_obj = synStore(obj,
                    used = ALL_USED_IDs, 
                    activityName = activityName, 
                    executed = thisFile, 
                    activityDescription = activityDescription)

# Raw Expression
write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file='MSSMlogCPM.tsv', sep = '\t', row.names=F, quote=F)
obj = File('MSSMlogCPM.tsv', 
           name = 'Raw logCPM',
           parentId = CODE$properties$id)
rawcpm_obj = synStore(obj,
                    used = ALL_USED_IDs, 
                    activityName = activityName, 
                    executed = thisFile, 
                    activityDescription = activityDescription)
